Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-04 examples.tex
+

to map new pages into an address space by writing into an existing page table (\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t pa, uint64\_t perm )} in Figure \ref{fig:mapping_code}).

@@

\newcommand{\fpaddr}{\texttt{fpaddr}}
\newcommand{\specline}[1]{{\color{blue}\left\{#1\right\}}}
\newcommand{\sumapacesfull}[2]{
  \ownGhost\gammaPreds{\authfull{\singletonMap{#1}{#2}}}
}
\todo{ismail change the following spec base on the changes you have in the code}
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
;;complstatus_t vaspace_mappage(pte_t *pml4, void *va,uintptr_t pa, uint64_t perm ) {
...
;;setting up the stack      
;;char *a; pte_t *pte;
;;a = (char *)PGROUNDUP((uintptr_t)va);
 $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,m)  \ast  \texttt{r14}\mapsto_{\textsf{r}} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{\textsf{r}} \_ \ast \ulcorner \texttt{aligned va} \land  \theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$
;;char *a; pte_t *pte;
;;a = (char *)PGROUNDUP((uintptr_t)va);
mov    -0x10[rbp],rcx
add    0x1000,rcx
sub    0x1,rcx
and    0xfffffffffffff000,%rcx
mov    rcx,-0x28[rbp]

;;pte = walkpgdir_succ(pml4, a);
mov    -0x8[rbp],rdi
mov    -0x28[rbp],rsi
mov    0x0,rax
callq  291 <vaspace_mappage+0x41>	28d: R_X86_64_PLT32	walkpgdir_succ-0x4
mov    0x0,r8
movslq eax,rcx
mov    rcx,-0x30[rbp]
;;pte->writable     = PTE_WRITEABLE_SET(perm);
...
pte->user_acc     = PTE_USER_ACC_SET(perm);
...
;;pte->writethrough = PTE_WRITETHRU_SET(perm);
...
;;pte->nocache      = PTE_NOCACHE_SET(perm);
pte->pfn          = PTE_ADDR_TO_PFN(pa);
mov    -0x18[rbp],rcx
shr    0xc,rcx
mov    -0x30[rbp],rdx
mov    [rdx],rsi
movabs 0xffffffffff,rdi
and    rdi,rcx
shl    0xc,rcx
movabs 0xfff0000000000fff,rdi
and    rdi,rsi
or     rcx,rsi
mov    rsi,[rdx]
;;pte->nx           = PTE_NX_SET(perm);
;;pte->present      = 1;
;;clean up the stack and return
...
;;}
\end{lstlisting}
  \caption{Specification and proof of code for mapping a new page.}
\label{fig:mapping_code}
\end{figure}

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline|pte\_walkpgdir| (discussed again in Section \ref{sec:traversing}).
\textsf{pte\_walkpgdir}'s postcondition contains virtual \emph{PTE} pointsto assertions ($\mapsto_{\textsf{vpte}}$) \todo{colin, these are no more plain physical pointstos}  both for ensuring partial page table walk reaching the
L1 entry (l1e) with the justification of higher levels of the page table exist, and for allowing access to the memory of the L1 entry via virtual address.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte,q}} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\sumapaces{\mathit{cr3val}}\delta) \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
This is also guaranteed ($ \ulcorner
\texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$).
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

@@

The crucial step is actually updating the L1 entry,
via the virtual address
(\textsf{pte\_addr}) known to translate to the appropriate physical address, in our example the L1
table entry address ($\textsf{addr\_L1}(\textsf{va, l1e})$), to hold the freshly
allocated physical page address (\textsf{fpaddr}) in Line X.
After this write, we have (Specification Line Y)
\[\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr}  \qquad\qquad (\textsf{by the PTE variant of Rule}~\TirNameStyle{WriteToVirtMemFromReg})\]

>>>>

to map new pages into an address space by writing into an existing page table via a call
\[\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t pa, uint64\_t perm )}\] in Figure \ref{fig:mapping_code}).

@@

\newcommand{\fpaddr}{\texttt{fpaddr}}
\newcommand{\specline}[1]{{\color{blue}\left\{#1\right\}}}
\newcommand{\sumapacesfull}[2]{
  \ownGhost\gammaPreds{\authfull{\singletonMap{#1}{#2}}}
}
\todo{ismail change the following spec base on the changes you have in the code}
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
;;complstatus_t vaspace_mappage(pte_t *pml4, void *va,uintptr_t pa, uint64_t perm ) {
...
;;setting up the stack      
;;char *a; pte_t *pte;
;;a = (char *)PGROUNDUP((uintptr_t)va);
      $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,m)  \ast \texttt{rdi}\mapsto_{r} \textsf{plm4} \ast \texttt{rsi}\mapsto_{r} \textsf{va} \ast  \texttt{rdx}\mapsto_{r} \textsf{pa} \ast \texttt{rcx}\mapsto_{r} \textsf{perm} \ast \texttt{rax}\mapsto_{\textsf{r}} \_ \ast \ulcorner \texttt{aligned va} \land  \theta \; !!\; \vaddr = \texttt{None}\urcorner\ast }_{\rtv}$
      $\specline{\texttt{rbp-8}\mapsto_{v} \textsf{plm4} \ast \texttt{rbp-16}\mapsto_{v} \textsf{va}  \ast \texttt{rbp-24}\mapsto_{v} \textsf{pa} \ast \texttt{rbp-32}\mapsto_{v} \textsf{perm} }_{\rtv}$
;;char *a; pte_t *pte;
;;a = (char *)PGROUNDUP((uintptr_t)va);
...
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,m) \ast  \texttt{r14}\mapsto_{\textsf{r}} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{\textsf{r}} \vaddr \;\ast \nfpointsto{\mask\vaddr\maskfour\rtv}{\mask\vaddr\maskfouroff\rtv}\entryf\qone\naddr \ast}_{\rtv}$ 
$\specline{  \nfpointsto{\mask\vaddr\maskthree\entryf}{\mask\vaddr\maskthreeoff\entryf}\entrytr\qtwo\naddr \ast \nfpointsto{\mask\vaddr\masktwo\entrytr}{\mask\vaddr\masktwooff\entrytr}\paddr\qthree\naddr \;\ast}_{\rtv}$
$\specline{\texttt{pte\_addr} \mapsto_{\texttt{vpte}} \paddr\; (\texttt{wzero 64})  \ast \ulcorner \texttt{entry\_present } \entryf \land \texttt{entry\_present } \entrytr \land  \texttt{entry\_present } \entrytw \urcorner}_{\rtv}$
$\specline{\exists \texttt{ fpaddr} \ldotp \ulcorner \texttt{aligned fpaddr} \urcorner \ast \texttt{rax}\mapsto_{\textsf{r}} \texttt{fpaddr+3} \ast \texttt{fpaddr} \mapsto_{\textsf{p}} (\texttt{wzero 64}) \ast \ulcorner \texttt{entry\_present (fpaddr+3)}\urcorner}_{\rtv}$      
;;pte = walkpgdir_succ(pml4, a);
mov    -0x8[rbp],rdi
mov    -0x28[rbp],rsi
mov    0x0,rax
callq  291 <vaspace_mappage+0x41>	28d: R_X86_64_PLT32	walkpgdir_succ-0x4
mov    0x0,r8
mov    rax,rcx
mov    rcx,-0x30[rbp]
;;pte->writable     = PTE_WRITEABLE_SET(perm);
...
pte->user_acc     = PTE_USER_ACC_SET(perm);
...
;;pte->writethrough = PTE_WRITETHRU_SET(perm);
...
;;pte->nocache      = PTE_NOCACHE_SET(perm);
pte->pfn          = PTE_ADDR_TO_PFN(pa);
mov    -0x18[rbp],rcx
shr    0xc,rcx
mov    -0x30[rbp],rdx
mov    [rdx],rsi
movabs 0xffffffffff,rdi
and    rdi,rcx
shl    0xc,rcx
movabs 0xfff0000000000fff,rdi
and    rdi,rsi
or     rcx,rsi
mov    rsi,[rdx]
;;pte->nx           = PTE_NX_SET(perm);
;;pte->present      = 1;
;;clean up the stack and return
...
;;}
\end{lstlisting}
  \caption{Specification and proof of code for mapping a new page.}
\label{fig:mapping_code}
\end{figure}

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline{pte\_walkpgdir} (discussed again in Section \ref{sec:traversing}).
\textsf{pte\_walkpgdir}'s postcondition contains virtual \emph{PTE} pointsto assertions ($\mapsto_{\textsf{vpte}}$) \todo{colin, these are no more plain physical pointstos}  both for ensuring partial page table walk reaching the
L1 entry (l1e) with the justification of higher levels of the page table exist, and for allowing access to the memory of the L1 entry via virtual address.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte,q}} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\sumapaces{\mathit{cr3val}}\delta) \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
This is also guaranteed ($ \ulcorner
\texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$).
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

@@

The crucial step is actually updating the L1 entry (Line X\todo{X} in Figure \ref{fig:mapping_code}),
via the virtual address
(\textsf{pte\_addr}) known to translate to the appropriate physical address, in our example the L1
table entry address ($\textsf{addr\_L1}(\textsf{va, l1e})$), to hold the freshly
allocated physical page address (\textsf{fpaddr}) in Line X.
After this write, we have (Specification Line Y)
\[\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr}  \qquad\qquad (\textsf{by the PTE variant of Rule}~\TirNameStyle{WriteToVirtMemFromReg})\]
